SMT-Based Generation of Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-03-17 2019-ActaInformatica/ActaInformatica.tex
more changes

@@

\section{Operational Semantics for Open pNets}
\label{section:op-semantics}

The semantics of open pNets will be defined as an open automaton, that
is an automaton where each transition composes transitions 
of several
% \noteSB{These should all be PLTSs, right? Or did I miss
%  something?}\noteEM{Hum, nor really, something intermediate; the actions there can be
%  parameterised actions, but the label has no guard and no
%  assignments}
LTSs with the actions of some holes; the transition occurs if some predicates hold,
and can involve a set of state modifications. Each state of an open
automaton has a set of \emph{state variables} that can be assigned by
incoming transitions. Strictly speaking, the LTSs at the leaves of the
open automaton are a restricted form of PLTSs, where labels are
parametrised actions, but include no guard nor assignments.

\begin{definition}[Open transitions]
	\label{def:OpenTransitions}
	An \emph{open transition} $OT$ over
        % a set  $\langle S_i,s_{0 i}, \rightarrow_i\rangle^{i\in I}$ of LTSs,
        a set $J$ of holes, %with sorts $Sort_j^{j\in J}$
        and a set of states $\mathcal{S}$ is a structure of the form:	
	\begin{mathpar}
	\inferrule*[myfraction=\reddottedrule]
	{%\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
		\{\xrightarrow{\beta_j}_j\}^{j\in J}, \Pred\,, \Post}
	{s \xrightarrow {\alpha}s'}\,,
	\end{mathpar}
	where $s, s'\in\mathcal{S}$, the $a_i, \beta_j, \alpha$ are
        action expressions,
        % $s_i{\xrightarrow{a_i}}_i s_i^{\prime}$ is a transition of the 
	% LTS $\langle S_i,s_{0 i}, \rightarrow_i\rangle$
        $\beta_j$
        is an action of the hole $j$, and $\alpha$ is the resulting action
        of $OT$. \Pred\ is a predicate
	over the variables of the terms, labels, and states $s_i$,
        $\beta_j$, $s$, $\alpha$. \Post\ is a set of equations that  
	hold \emph{after the open transition}, represented as a
        substitution $\{x_k\gets e_k\}^{k\in K}$  
	where $x_k$ are variables of $s'$ and $s'_i$, whereas $e_k$ are
        expressions over the other variables of the open transition.
\end{definition}

\begin{definition}[Open automaton]
	\label{def:open-automaton}
	An \emph{open automaton} is a structure\\ $A =
	\langle J,\mathcal{S},s_0,\mathcal{T}\rangle$ where:
	\begin{itemize}
		\item[$\bullet$]  $I$ and $J$ are  sets of indices,
	%	\item[$\bullet$]  $LTS_i^{i\in I}$ is a family of LTSs,
		\item[$\bullet$]  $\mathcal{S}$ is a set of states and $s_0 \in \mathcal{S}$ the initial state,
		\item[$\bullet$] $\mathcal{T}$ is a set of open transitions and, for each
		$t\in \mathcal{T}$, there exist $I'$, $J'$ with $I'\subseteq I$, $J'
		\subseteq J$, such that $t$ is an open transition over $J'$,
		and  $\mathcal{S}$.

@@

\paragraph{States of open pNets.}\label{def-states}
  A state of an open pNet is a tuple of the
  states of its leaves (in which we denote tuples
  in structured states as $\triangleleft\ldots\triangleright$).
  For any pNet p, let $\overline{\mathit{Leaves}} = \mylangle S_i,{s_i}_0, \to_i\myrangle^{i \in L}$ be the set of PLTS at its leaves,
  then $\mathit{States}(p) = \{\ostate{s_i^{i\in L}}
  |\, \forall i\in L. s_i \in S_i\}$.
A PLTS being its own single leave:
  $\mathit{States}(\mylangle S,s_0, \to\myrangle) = \{\ostate{s}|\, s \in S\}$.
The initial state is defined as:
$\mathit{InitState}(p) = \ostate{{{s_i}_0}^{i\in L}}$.

@@

\begin{example}[An open transition]\TODO{change for Enable example}
  Transition $OT_1$ of the pNet $EnableCompl$ is generated by
  combination of the first vector of its root node, with
  the second vector of its $2^{nd}$ node. Its generated predicate is
  trivialy true. The global states shows that controler $C_3$ has
    changed from state 0 to 1.

@@

In \cite{henrio:Forte2016} we also proved:
\begin{theorem}[Finiteness]
  \label{Th:finiteness}
  Let $pnet=\mylangle \overline{\pNet},\overline{S}, \symb{SV}_k^{k\in K}\myrangle$ be an open pNet  with leaves $l_i^{i\in I}$ and holes $h_j^{j\in
  J}$, where the sets $I$ and $J$ are finite. Suppose the synchronisation vectors of all pNets 
  included in  $pnet$   are finite, and 
% $\forall i \in L.\, finite{(states(l_i))} \text{ and } l_i$ 
$l_i $ has a finite number of state variables for each $i\in I$, then the semantics of $pnet$ is an open automaton 
%$\mathcal{T}$ 
with finitely many states and transitions.
\end{theorem}
  Remark that all the elements of such pNets and open automata being
symbolic, they can represent many classes of unbounded systems.

@@

The inside loop (\emph{MakeTransitions} method) applies recursively
the semantic rules following the structure of the pNet.
When applied to a PLTSs at the leaves, we simply take the PLTS transitions of the
corresponding local state and use the semantic rule to build the OT\footnote{%
%
  We omit detailed presentation of this case for the sake of brevity.
%
}.
When applied to a pNet node we use two methods, combining and
matching, to generate the open transitions in a hierarchical manner,
as shown in Alg.~\ref{alg2}. This method directly manages the
holes of the node, so \emph{MakeTransitions} is never called on a hole.

@@

\def\inactive{\{-\}}
\paragraph{Combining.}
The combining method enumerates all the possible behaviours of the
subnets as all the possible combinations of their open transitions.
Assume that there is a collection of $n$ subnets.
%% \noteSB{These variables are not used and conflict with the returned
%%   set $L$ in Alg.~\ref{alg2}, on one hand, and control states $s_i$ of
%%   the subnets, on the other.}
%% \st{$L = {s_1, s_2, ...,  s_n}$;}
We denote $\overline{ot}_i$ the set of open
transitions of the $i$-th
subnet (obtained in line~3 of the algorithm); %% $s_i$ and 
``$-$'' means that the subnet is not involved.
The combination $\overline{comb}$, a set of n-tuples, is the
cartesian product:  
$\overline{comb} = (\inactive\cup \overline{ot}_1) \times (\inactive\cup \overline{ot}_2)\times \dots \times (\inactive\cup \overline{ot}_n)$.

@@

\paragraph{Matching.}
The \emph{Matching} method builds the OTs of a pNet node from those of
its subnets.
For each synchronisation vector and each possible
combination of behaviours of the subnets, as generated by the \emph
{Combining} method, it builds the corresponding open transition.
Here, we only detail the construction of the predicate.
Suppose $sv = \left((a'_i)^{i\in I} (b'_j)^{j\in J}\rightarrow v'\right) \in SV$ is a synchronisation vector, $G_k$ is its guard, 
$C = ({ot}_i)^{i \in [1,n]} \in \overline{comb}$ is a tuple of open transitions such that, for each $i \in [1,n]$, either $ot_i = -$, or the result action of $ot_i$ is $a_i$. Suppose futher that $\mathit{hole} = (b_j)^{j \in J}$ is the hole behaviour and $v$ is a fresh variable, representing the result action of the OT under construction. Then we can  build the predicate:
%
\begin{multline*}
  \mathit{MkPred}(sv, C, \overline{\mathit{hole}}) = 
  \\
  (\forall i\in I, a_i=a'_i) \land
%  (\forall i \not\in I, ot_i = -) \land
  (\forall j\in J, b_j=b'_j) \land
  (v=v') \land G_k.
\end{multline*}

@@

Checking satisfiability requires some symbolic computation
on the action expressions and the predicates, which
may depend on the specific theory of the action algebra datatypes. 
The ``Modulo Theory'' part of SMT solvers is important here, as
the solver can use specific properties of each action algebra.

@@

Algorithm~\ref{alg3} implement this new stategy: \TODO{Explain
  significant detials}

@@

Our implementation submits satisfiability requests to
Z3 using its JAVA API. Here, for readability, we
show the Z3 code using its SMT-LIB input language.
Note that in the previous sections, the OTs were displayed in a
  simplified, human readable form. The input and output of our tool, and
  also the generated SMTlib fragments, are
  slightly more difficult to read, in 
particular because of structured names for the fresh variables generated
by the algorithm, allowing tracability of the result~\cite{Avocs-RR}.

Figure~\ref{schema:smt-lib} shows the translation of the
transition of Figure~\ref{schema:unsat-ot}
in the SMTlib syntax.
It contains the declaration
of the BIP action algebra sorts 
and constructors, then the declaration of variables, and finally the
predicate to be checked, encoded as a set of assertions.
Here the behavior of the hole ``B'' appears as a variable
``|hb\_B:13:1|'', and line 10 shows the constraint that matching with
a synchronisation vector has imposed on the behavior of ``B''.
Here we display also the diagnosis (``sat'' or ``unsat'') generated by Z3.

@@

For simplicity, we have only shown here a use-case with boolean data
parameters, so we only require the basic SAT capability of the Z3
engine. It should be clear that more involved examples will require
reasonning about predicates of various data types. Z3 provides theories for reasonning
about integers, bitstrings, enumerated types, arrays, etc.
We also need to deal with some universal quantification on free
variables occuring in the guards of synchronisation vectors.
For other data structures  in such models, like records, or
recursive structures, the user will have to develop its own theories.
Decidability of these theories will condition the completeness of the
satisfiability check, and the decidability of model-checking or
bisimulation checking.

@@

\emph{Translation of action algebra presentation.}
In~\cite{Avocs-RR}, we  define the translation of an algebra presentation into
SMTlib declarations (\texttt{declare-datatypes} and
\texttt{declare-fun}). We also formalise a condition of well-formedness of pNets to ensure that the generated code is correct,
and will not raise runtime errors in the SMT engine. Note that the
\texttt{declare-datatypes} command comprises both the action
constructors from Table~\ref{table:BIPalgebra} and also the constant action
names from our example.
Additionally, we will include axiomatization of any required functions
and predicates of the presentation data-types.

@@

Application of the Failure Monitor architecture ensures that, whenever
a failure is registered in the operand component, the system will be
reset, unless a resumption is registered within $\mathrm{Max}$ time
units (more details in Sect.~\ref{section:resultOA}).

@@

Figure~\ref{schema:ArchFailure:pNet} shows a pNet encoding of the
above Failure Monitor architecture.  This encoding is structural: each
coordinator component is encoded as a PLTS, the operand
component\mdash as a hole; connectors of the BIP model are encoded as
synchronisation vectors.  Each connector that does not involve
triggers is trivially encoded by a synchronisation vector comprising
the same ports.  In order to encode the semantics of the connectors
involving triggers, we 1)~in the PLTS encoding the coordinator
components, add loop transitions to ensure that all ports involved in
such connectors are enabled in all states, 2)~associate a Boolean
value to each of these ports: the original transitions carry the value
$\true$ (\eg $\cstate{0} \goesto{\PortFail(\true)} \cstate{1}$) and the
added loops carry the value $\false$ (\eg $\cstate{2}
\goesto{\PortFail(\false)} \cstate{2}$), 3)~add to the corresponding
synchronisation vectors the Boolean predicate encoding the connector
structure.  For example, $SV0$ encodes the connector discussed above:
the predicate $(b1=b2) \land (b1\lor b2 \Rightarrow b0)$ means that the
``true'' transitions $\NameCtrl.\PortFail$ and $\NameTimer.\PortStart$
can only fire together ($b1 = b2$) and whenever one of them fires,
$\NameOprnd.\PortFail$ must fire also ($b1 \lor b2 \Rightarrow b0$).
This encoding can be systematically obtained for any hierarchical BIP
connector~\cite{BliSif10-causal-fmsd}.
%
Although, for the sake of brevity, we omit priorities from the
encoding, this can be easily achieved, by introducing additional
Boolean variables for relevant ports~\cite{BarBliu15-offer-scico}.

@@

Figure \ref{fig:ArchFailure:OA} shows the open automaton built by our
algorithm. For reasons of readability, we do not show the full
details of the open transitions on this figure. Below we list the
first three of them (full details available in \ref{TimerRR}). 
This automaton has 12 transitions, including those encoding various
possible firing of some interactions (e.g. OTs 2 and 3 for
{\PortFail}). It appears that the rough behaviour is
reasonable, although a detailed study shows that the property stated
above does not hold, because of the
{\PortFail} loops on states $11$ and $12$. The reason is that we do not
use the maximal progress assumption here. If we do, we get the
corrected behaviour in Figure \ref{fig:ArchFailure:OA-MaxProgress}, in
which $OT_7$ has disappeared, and $OT_2$ and 
$OT_{10}$ have been restricted to $b0=\false$.

@@

The formal definitions and properties of the open pNets model were
given in \cite{henrio:Forte2016}. In the current work we describe an implementation of
the model and its semantics construction, including its interaction
with the Z3 SMT engine.
The implementation has two parts: the first is an algorithm
that builds all possible combinations of synchronisations through the
pNet hierarchical structure. The result is a so-called
\emph{open-automaton}, whose transitions contain predicates relating
the actions of the pNet holes and controllers. Some of the open
transitions obtained at this step  may
contain predicates which do not represent any possible concrete
instantiations. 
In the second part of the tool we use the SMT solver Z3 to check the
satisfiability of the predicate in each open transition. 
To this end, we encode into Z3 the representations of the action algebra and
the predicates before submitting them to the
Z3 solver. 
%
%\hl{%
%
In this paper, we used a running example, based on a BIP
architecture from an earlier nanosatellite case
study~\cite{CubETH-case-study}.  This example shows that
open-automata-based semantics can be instrumental in verifying the
properties enforced by the architectures through an encoding into open
pNets.  This encoding %\mdash which we intend to formalise and prove
% correct in a separate paper\mdash 
also opens the way for an extension
of BIP architectures with the transfer of data among variables of
different components.  Indeed, such data transfer can be easily
encoded using the predicates associated to synchronisation vectors in
open pNets.  The encoding of open transitions into SMTlib and the
availability of theories can guide the definition of such an extension.
%
%}
%
Our case-studies show that our encoding
successfully identifies the unsatisfiable open transitions and that
the resulting automata correctly reflect the expected 
movements of the encoded process expressions.

@@

%% \todoMargin{Change perspective: we are working on the formal extension of
%%   BIP architectures, and their semantics in terms of pNets. Then
%%   bisimulation and model-checking principles adapted to open pNet
%%   systems are under development}
%
Naturally, our next goals after the generation of the open automata
will be to model-check logical properties, and to check equivalences of pNets. While
model-checking open automata  seems easy to define, equivalence
checking is more challenging. 
In~\cite{henrio:Forte2016}, we
have already found the FH-bisimulation, %inspired
%by~\cite{deSimone85},
to be a
suitable definition. But weak equivalences, or refinements, will
definitely be useful when comparing different pNets with
different structure. For bisimulation, we foresee that SMT methods
will be the basis for comparing open transitions. 
%However, the usual approach of computing bisimulation by partition
%% refinement does not work easily with symbolic transitions. In a first
%% step we may want instead to define and implement an algorithm for
%% checking that a given equivalence relation between open automata 
%% states is indeed a FH-Bisimulation. Finding automatically a
%% bisimulation relation or the most general bisimulation will be more
%% challenging.
%% It appears also that when comparing different pNets with
%% different structure, strong bisimulation will be too strong a notion,
%% and that we will have to define proper notions of weak equivalence or
%% refinements.

>>>>

\renewcommand{\pLTS}{PLTS}

@@

\section{Operational Semantics of Open pNets}
\label{section:op-semantics}

The semantics of open pNets will be defined as an open automaton, that
is an automaton where each transition composes transitions 
of several
% \noteSB{These should all be PLTSs, right? Or did I miss
%  something?}\noteEM{Hum, nor really, something intermediate; the actions there can be
%  parameterised actions, but the label has no guard and no
%  assignments}
LTSs with the actions of some holes; the transition occurs if some predicates hold,
and can involve a set of state modifications. Each state of an open
automaton has a set of \emph{state variables} that can be assigned by
incoming transitions. Strictly speaking, the LTSs at the leaves of the
open automaton are a restricted form of PLTSs, where labels are
parametrised actions, but include neither guard nor assignments.

\begin{definition}[Open transitions]
	\label{def:OpenTransitions}
	An \emph{open transition} $OT$ over
        % a set  $\langle S_i,s_{0 i}, \rightarrow_i\rangle^{i\in I}$ of LTSs,
        a set $J$ of holes %with sorts $Sort_j^{j\in J}$
        and a set of states $\mathcal{S}$ is a structure of the form:	
	\begin{mathpar}
	\inferrule*[myfraction=\reddottedrule]
	{%\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
		\{\xrightarrow{\beta_j}_j\}^{j\in J}, \Pred\,, \Post}
	{s \xrightarrow {\alpha}s'}\,,
	\end{mathpar}
	where $s, s'\in\mathcal{S}$, the $a_i, \beta_j, \alpha$ are
        action expressions,
        % $s_i{\xrightarrow{a_i}}_i s_i^{\prime}$ is a transition of the 
	% LTS $\langle S_i,s_{0 i}, \rightarrow_i\rangle$
        $\beta_j$
        is an action of the hole $j$, and $\alpha$ is the resulting action
        of $OT$. \Pred\ is a predicate
	over the variables of the terms, labels, and states $s_i$,
        $\beta_j$, $s$, $\alpha$. \Post\ is a set of equations that  
	hold \emph{after the open transition}, represented as a
        substitution $\{x_k\gets e_k\}^{k\in K}$  
	where $x_k$ is a variable of $s'$ or $s'_i$, whereas $e_k$ is an
        expression over other variables of the open transition.
\end{definition}

\begin{definition}[Open automaton]
	\label{def:open-automaton}
	An \emph{open automaton} is a structure\\ $A =
	\langle J,\mathcal{S},s_0,\mathcal{T}\rangle$ where:
	\begin{itemize}
		\item[$\bullet$]  $I$ and $J$ are  sets of indices,
	%	\item[$\bullet$]  $LTS_i^{i\in I}$ is a family of LTSs,
		\item[$\bullet$]  $\mathcal{S}$ is a set of states and $s_0 \in \mathcal{S}$ the initial state,
		\item[$\bullet$] $\mathcal{T}$ is a set of open transitions and, for each
		$t\in \mathcal{T}$, there exist $I'$, $J'$ with $I'\subseteq I$, $J'
		\subseteq J$, such that $t$ is an open transition over $J'$
		and  $\mathcal{S}$.

@@

\paragraph{States of open pNets.}\label{def-states}
  A state of an open pNet is a tuple of the
  states of its leaves (in which we denote tuples
  in structured states as $\triangleleft\ldots\triangleright$).
  For any pNet p, let $\overline{\mathit{Leaves}} = \mylangle S_i,{s_i}_0, \to_i\myrangle^{i \in L}$ be the set of PLTSs at its leaves,
  then $\mathit{States}(p) = \{\ostate{s_i^{i\in L}}
  |\, \forall i\in L. s_i \in S_i\}$.
A PLTS could be its own single leave:
  $\mathit{States}(\mylangle S,s_0, \to\myrangle) = \{\ostate{s}|\, s \in S\}$.
The initial state is defined as:
$\mathit{InitState}(p) = \ostate{{{s_i}_0}^{i\in L}}$.

@@

\begin{example}[An open transition]\TODO{change for Enable example}
  Transition $OT_1$ of the pNet $EnableCompl$ is generated by
  combining the first vector of its root node with
  the second vector of its second node. Its generated predicate is
  trivially true. The global state shows that controler $C_3$ has
    changed from state 0 to 1.

@@

In \cite{henrio:Forte2016} we also proved:
\begin{theorem}[Finiteness]
  \label{Th:finiteness}
  Let $pnet=\mylangle \overline{\pNet},\overline{S}, \symb{SV}_k^{k\in K}\myrangle$ be an open pNet  with leaves $l_i^{i\in I}$ and holes $h_j^{j\in
  J}$, where the sets $I$ and $J$ are finite. Suppose the synchronisation vectors of all pNets 
  included in  $pnet$   are finite, and 
% $\forall i \in L.\, finite{(states(l_i))} \text{ and } l_i$ 
$l_i $ has a finite number of state variables for each $i\in I$. Then the semantics of $pnet$ is an open automaton 
%$\mathcal{T}$ 
with finitely many states and transitions.
\end{theorem}
  Remark that all the elements of such pNets and open automata being
symbolic, they can represent many classes of unbounded systems.

@@

The inside loop (\emph{MakeTransitions} method) applies recursively
the semantic rules following the structure of the pNet.
When applying a rule to a PLTS at the leaves, we simply take the PLTS transitions of the
corresponding local state and use the semantic rule to build the OT\footnote{%
%
  We omit detailed presentation of this case for the sake of brevity.
%
}.
When applying a rule to a pNet node we use two methods, combining and
matching, to generate the open transitions in a hierarchical manner,
as shown in Algorithm~\ref{alg2}. This method directly manages the
holes of the node, so \emph{MakeTransitions} is never called on a hole.

@@

\def\inactive{\{-\}}
\paragraph{Combining.}
The combining method enumerates all the possible behaviours of the
subnets as all the possible combinations of their open transitions.
Assume that there is a collection of $n$ subnets.
%% \noteSB{These variables are not used and conflict with the returned
%%   set $L$ in Alg.~\ref{alg2}, on one hand, and control states $s_i$ of
%%   the subnets, on the other.}
%% \st{$L = {s_1, s_2, ...,  s_n}$;}
We denote by $\overline{ot}_i$ the set of open
transitions of the $i$-th
subnet (obtained in line~3 of the algorithm); %% $s_i$ and 
``$-$'' means that the subnet is not involved.
The combination $\overline{comb}$, a set of $n$-tuples, is the
cartesian product:  
$\overline{comb} = (\inactive\cup \overline{ot}_1) \times (\inactive\cup \overline{ot}_2)\times \dots \times (\inactive\cup \overline{ot}_n)$.

@@

\paragraph{Matching.}
The \emph{Matching} method builds the OTs of a pNet node from those of
its subnets.
For each synchronisation vector and each possible
combination of behaviours of the subnets, as generated by the \emph
{Combining} method, it builds the corresponding open transition.
Here, we only detail the construction of the predicate.
Suppose that $sv = \left((a'_i)^{i\in I} (b'_j)^{j\in J}\rightarrow v'\right) \in SV$ is a synchronisation vector, $G_k$ is its guard, and
$C = ({ot}_i)^{i \in [1,n]} \in \overline{comb}$ is a tuple of open transitions such that, for each $i \in [1,n]$, either $ot_i = -$, or the result action of $ot_i$ is $a_i$. Suppose futher that $\mathit{hole} = (b_j)^{j \in J}$ is the hole behaviour and $v$ is a fresh variable, representing the result action of the OT under construction. Then we can  build the predicate:
%
\begin{multline*}
  \mathit{MkPred}(sv, C, \overline{\mathit{hole}}) = 
  \\
  (\forall i\in I, a_i=a'_i) \land
%  (\forall i \not\in I, ot_i = -) \land
  (\forall j\in J, b_j=b'_j) \land
  (v=v') \land G_k.
            \end{multline*}

@@

Checking satisfiability requires some symbolic computation
on the action expressions and the predicates, which
may depend on the specific theory of the action algebra datatypes. 
The ``Modulo Theory'' part of SMT solvers is important here, as
the solvers can use specific properties of each action algebra.

@@

Algorithm~\ref{alg3} implements this new stategy: \TODO{Explain
  significant detials}

@@

Our implementation submits satisfiability requests to
Z3 using its Java API. Here, for readability, we
show the Z3 code using its SMT-LIB input language.
Note that in the previous sections, the OTs were displayed in a
  simplified, human readable form. The input and output of our tool, and
  also the generated SMTlib fragments, are
  slightly more difficult to read, in 
particular because of structured names for the fresh variables generated
by the algorithm, allowing for tracability of the result~\cite{Avocs-RR}.

Figure~\ref{schema:smt-lib} shows the translation of  Figure~\ref{schema:unsat-ot}
in the SMTlib syntax.
It contains the declaration
of the BIP action algebra sorts 
and constructors, then the declaration of variables, and finally the
predicate to be checked, encoded as a set of assertions.
Here the behavior of the hole ``B'' appears as a variable
``|hb\_B:13:1|'', and line 10 shows the constraint that matching with
a synchronisation vector has imposed on the behavior of ``B''.
Here we display also the diagnosis (``sat'' or ``unsat'') generated by Z3.

@@

For simplicity, we have only shown here a use-case with boolean data
parameters, so we only require the basic SAT capability of the Z3
engine. It should be clear that more involved examples will require
reasoning about predicates of various data types. Z3 provides theories for reasonning
about integers, bitstrings, enumerated types, arrays, etc.
We also need to deal with some universal quantification on free
variables occuring in the guards of synchronisation vectors.
For other data structures  in such models, like records, or
recursive structures, the user will have to develop its own theories.
Decidability of these theories will condition on the completeness of the
satisfiability check, and the decidability of model-checking or
bisimulation checking.

@@

\emph{Translation of action algebra presentation.}
In~\cite{Avocs-RR}, we  define the translation of an algebra presentation into
SMTlib declarations (\texttt{declare-datatypes} and
\texttt{declare-fun}). We also formalise a condition of well-formedness of pNets to ensure that the generated code is correct,
and will not raise runtime errors in the SMT engine. Note that the
\texttt{declare-datatypes} command comprises both the action
constructors from Table~\ref{table:BIPalgebra} and also the constant action
names from our example.
Additionally, we will include axiomatisation of any required functions
and predicates of the presentation data-types.

@@

Application of the Failure Monitor architecture ensures that, whenever
a failure is registered in the operand component, the system will be
reset, unless a resumption is registered within $\mathrm{Max}$ time
units (more details in Section~\ref{section:resultOA}).

@@

Figure~\ref{schema:ArchFailure:pNet} shows a pNet encoding of the
above Failure Monitor architecture.  This encoding is structural: each
coordinator component is encoded as a PLTS, the operand
component as a hole; connectors of the BIP model are encoded as
synchronisation vectors.  Each connector that does not involve
triggers is trivially encoded by a synchronisation vector comprising
the same ports.  In order to encode the semantics of the connectors
involving triggers, we 1)~in the PLTS encoding the coordinator
components, add loop transitions to ensure that all ports involved in
such connectors are enabled in all states, 2)~associate a Boolean
value to each of these ports: the original transitions carry the value
$\true$ (\eg $\cstate{0} \goesto{\PortFail(\true)} \cstate{1}$) and the
added loops carry the value $\false$ (\eg $\cstate{2}
\goesto{\PortFail(\false)} \cstate{2}$), 3)~add to the corresponding
synchronisation vectors the Boolean predicate encoding the connector
structure.  For example, $SV0$ encodes the connector discussed above:
the predicate $(b1=b2) \land (b1\lor b2 \Rightarrow b0)$ means that the
``true'' transitions $\NameCtrl.\PortFail$ and $\NameTimer.\PortStart$
can only fire together ($b1 = b2$) and whenever one of them fires,
$\NameOprnd.\PortFail$ must fire also ($b1 \lor b2 \Rightarrow b0$).
This encoding can be systematically obtained for any hierarchical BIP
connector~\cite{BliSif10-causal-fmsd}.
%
Although, for the sake of brevity, we omit priorities from the
encoding, this can be easily achieved, by introducing additional
Boolean variables for relevant ports~\cite{BarBliu15-offer-scico}.

@@

Figure \ref{fig:ArchFailure:OA} shows the open automaton built by our
algorithm. For reasons of readability, we do not show the full
details of the open transitions in this figure. Below we list the
first three of them (full details available in \ref{TimerRR}). 
This automaton has 12 transitions, including those encoding various
possible firing of some interactions (e.g. OTs 2 and 3 for
{\PortFail}). It appears that the rough behaviour is
reasonable, although a detailed study shows that the property stated
above does not hold, because of the
{\PortFail} loops on states $11$ and $12$. The reason is that we do not
use the maximal progress assumption here. If we do, we get the
corrected behaviour in Figure \ref{fig:ArchFailure:OA-MaxProgress}, in
which $OT_7$ has disappeared, and $OT_2$ and 
$OT_{10}$ have been restricted to $b0=\false$.

@@

The formal definitions and properties of the open pNets model were
given in \cite{henrio:Forte2016}. In the current work we describe an implementation of
the model and its semantics construction, including its interaction
with the Z3 SMT engine.
The implementation has two parts: the first is an algorithm
that builds all possible combinations of synchronisations through the
pNet hierarchical structure. The result is a so-called
\emph{open automaton}, whose transitions contain predicates relating
the actions of the pNet holes and controllers. Some of the open
transitions obtained at this step  may
contain predicates which do not represent any possible concrete
instantiations. 
In the second part of the tool we use the SMT solver Z3 to check the
satisfiability of the predicate in each open transition. 
To this end, we encode into Z3 the representations of the action algebra and
the predicates before submitting them to the
Z3 solver. 
%
%\hl{%
%
In this paper, we used a running example, based on a BIP
architecture from an earlier nanosatellite case
study~\cite{CubETH-case-study}.  This example shows that
open-automata-based semantics can be instrumental in verifying the
properties enforced by the architectures through an encoding into open
pNets.  This encoding %\mdash which we intend to formalise and prove
% correct in a separate paper\mdash 
also opens the way for an extension
of BIP architectures with the transfer of data among variables of
different components.  Indeed, such data transfer can be easily
encoded using the predicates associated to synchronisation vectors in
open pNets.  The encoding of open transitions into SMTlib and the
availability of theories can guide the definition of such an extension.
%
%}
%
Our case-studies show that our encoding
successfully identifies the unsatisfiable open transitions and that
the resulting automata correctly reflect the expected 
movements of the encoded process expressions.

@@

%% \todoMargin{Change perspective: we are working on the formal extension of
%%   BIP architectures, and their semantics in terms of pNets. Then
%%   bisimulation and model-checking principles adapted to open pNet
%%   systems are under development}
%
Naturally, our next goals after the generation of the open automata
will be to model-check logical properties, and to check equivalences of pNets. While
model-checking open automata  seems easy to define, equivalence
checking is more challenging. 
In~\cite{henrio:Forte2016}, we
have already found the FH-bisimulation %inspired
%by~\cite{deSimone85},
to be a
suitable definition. But weak equivalences, or refinements, will
definitely be useful when comparing different pNets with
different structure. For bisimulation, we foresee that SMT methods
will be the basis for comparing open transitions. 
%However, the usual approach of computing bisimulation by partition
%% refinement does not work easily with symbolic transitions. In a first
%% step we may want instead to define and implement an algorithm for
%% checking that a given equivalence relation between open automata 
%% states is indeed a FH-Bisimulation. Finding automatically a
%% bisimulation relation or the most general bisimulation will be more
%% challenging.
%% It appears also that when comparing different pNets with
%% different structure, strong bisimulation will be too strong a notion,
%% and that we will have to define proper notions of weak equivalence or
%% refinements.
